Evolution of pre-copulatory and post-copulatory inbreeding behaviour

### 2016-09-22 ms_pre_post.Rmd
Fix last paragraph of the Introduction

<!--- I fear we're starting down a rabbit hole here. ---> 
Severe inbreeding depression might cause selection for alleles underlying active pre- and post-copulatory inbreeding avoidance and associated polyandry, but if evolution of one strategy renders its complement redundant for avoiding inbreeding, then selection might instead be sustained over generations for only one strategy. Empirical studies of multiple species have found evidence for female pre-copulatory inbreeding avoidance [e.g., @Lihoreau2007; @Kuriwada2011; @Kingma2013; @Fischer2015], and for post-copulatory inbreeding avoidance [e.g., @Simmons2006; @Bretman2009]. Further, evidence of both pre-copulatory and post-copulatory inbreeding avoidance has been found for a handful of well-studied species, including Trinidadian guppies [*Poecilia reticulata*; @Gasparini2011; @Daniel2015] and house mice [*Mus domesticus*; @Potts1991; @Firman2015], while @Liu2014 found evidence for pre-copulatory, but not post-copulatory, inbreeding avoidance in an experimental study of cabbage beetles (*Colaphellus bowringi*). These empirical studies demonstrate that all combinations of adaptive pre-copulatory and post-copulatory inbreeding avoidance are possible, but no conceptual framework exists to predict *a priori* what combinations should be favoured by selection when both pre- and post-copulatory inbreeding avoidance, and polyandry, can evolve. Frequencies of alleles underlying active pre- and post-copulatory inbreeding avoidance and polyandry are likely to be affected by direct negative selection on phenotypes (i.e., costs). For example, females that are genetically predisposed to polyandry often pay a cost associated with searching or making themselves available to mates, thereby increasing their own risk of predation [e.g., @Rowe1988; @Ronkainen1994; @Koga1998]. Females that express pre-copulatory choice (e.g., reluctance to mate with relatives) might pay a cost of increased risk of mating failure, or harm caused by sexual conflict [@Rowe1994; @Kokko2013]. Females that express post-copulatory choice might pay an energetic cost associated with physiological or biochemical mechanisms needed to store sperm and successfully bias fertilisation [e.g., @Gasparini2011; @Tuni2013; @Fitzpatrick2014b]. If the relative costs affecting pre- and post-copulatory inbreeding avoidance differ, then alleles underlying the less costly inbreeding strategy might become fixed in the population over generations, while alleles underlying the more costly inbreeding strategy might become extinct if their effects are both costly and made redundant by the less costly strategy.

In addition to the relative importance of costs, evolution of pre- or post-copulatory inbreeding avoidance might be precluded if inbreeding avoidance is already acheived through a complementary strategy. Previous models considered the fate of a rare allele underlying inbreeding avoidance in a population initially fixed for random mating [e.g., Duthie and Reid *In press*; @Parker1979; @Parker2006; @Duthie2016]. Such models thereby isolate the invasion fitness of a single focal strategy, but when multiple strategies affect realised inbreeding, it cannot be assumed that both will invade simultaneously, nor that invasion fitness of one inbreeding strategy will be independent of another. For example, if pre-adaptation or rapid selection resulted in the fixation of alleles underlying adaptive pre-copulatory inbreeding avoidance, then alleles underlying post-copulatory inbreeding avoidance might be precluded from invading a population in which post-copulatory choice is not already expressed because the effect that such invading alleles had on decreasing total inbreeding avoidance would be weak or negligible. To predict evolutionary outcomes, it is therefore necessary to consider evolution when one inbreeding avoidance strategy is initially rare, and the other strategy is common.

@@

Here we apply a model of female and male reproductive interactions and associated inbreeding depression in offspring viability to quantify evolution of three separate phenotypes including female tendency for polyandry, female pre-copulatory inbreeding strategy (mating strategy), and female post-copulatory inbreeding strategy (fertilisation strategy). We use individual-based modelling and *in silico* experiments to test the hypothesis that, given severe inbreeding depression in offspring viability, evolution of inbreeding avoidance through fertilisation bias causes selection for evolution of polyandry. Additionally, we test the hypothesis that costs associated with polyandry and inbreeding strategies will strongly affect evolution of each such that inbreeding avoidance through mate choice but not fertilisation bias will evolve when inbreeding mating strategy is less costly than the combined cost of inbreeding fertilisation strategy and tendency for polyandry, while inbreeding avoidance through fertilisation bias but not mate choice will evolve when inbreeding mating strategy is more costly than the combined costs of inbreeding fertilisation strategy and tendency for polyandry. Finally, we test the hypothesis that evolution of one inbreeding strategy will be precluded if it is introduced into a population that is already fixed for a complementary inbreeding strategy that is adaptive. We thereby provide a framework for predicting evolution of pre-copulatory and post-copulatory inbreeding avoidance across reproductive systems with different relative costs and strong inbreeding depression in offspring viability.

>>>>

<!--- I fear we're starting down a rabbit hole here. ---> 
Severe inbreeding depression might cause selection for alleles underlying active pre- and post-copulatory inbreeding avoidance and associated polyandry, but if evolution of one strategy renders its complement redundant for avoiding inbreeding, then selection might instead be sustained over generations for only one strategy. Empirical studies of multiple species have found evidence for female pre-copulatory inbreeding avoidance [e.g., @Lihoreau2007; @Kuriwada2011; @Kingma2013; @Fischer2015], and for post-copulatory inbreeding avoidance [e.g., @Simmons2006; @Bretman2009]. Further, evidence of both pre-copulatory and post-copulatory inbreeding avoidance has been found for a handful of well-studied species, including Trinidadian guppies [*Poecilia reticulata*; @Gasparini2011; @Daniel2015] and house mice [*Mus domesticus*; @Potts1991; @Firman2015], while @Liu2014 found evidence for pre-copulatory, but not post-copulatory, inbreeding avoidance in an experimental study of cabbage beetles (*Colaphellus bowringi*). These empirical studies demonstrate that all combinations of adaptive pre-copulatory and post-copulatory inbreeding avoidance are possible, but no conceptual framework exists to predict *a priori* what combinations should be favoured by selection when both pre- and post-copulatory inbreeding avoidance, and polyandry, can evolve.

Frequencies of alleles underlying active pre- and post-copulatory inbreeding avoidance and polyandry are likely to be affected by direct negative selection on phenotypes (i.e., costs). For example, females that are genetically predisposed to polyandry often pay a cost associated with searching or making themselves available to mates, thereby increasing their own risk of predation [e.g., @Rowe1988; @Ronkainen1994; @Koga1998]. Females that express pre-copulatory choice (e.g., reluctance to mate with relatives) might pay a cost of increased risk of mating failure, or harm caused by sexual conflict [@Rowe1994; @Kokko2013]. Females that express post-copulatory choice might pay an energetic cost associated with physiological or biochemical mechanisms needed to store sperm and successfully bias fertilisation [e.g., @Gasparini2011; @Tuni2013; @Fitzpatrick2014b]. If the relative costs affecting pre- and post-copulatory inbreeding avoidance differ, then alleles underlying the less costly inbreeding strategy might become fixed in the population over generations, while alleles underlying the more costly inbreeding strategy might become extinct if their effects are both costly and made redundant by the less costly strategy.

In addition to the relative importance of costs, evolution of pre- or post-copulatory inbreeding avoidance might be precluded if inbreeding avoidance is already acheived through a complementary strategy. Previous models considered the fate of a rare allele underlying inbreeding avoidance in a population initially fixed for random mating [e.g., Duthie and Reid *In press*; @Parker1979; @Parker2006; @Duthie]. Such models thereby isolate the invasion fitness of a single focal strategy, but when multiple distinct strategies affect realised inbreeding, it cannot be assumed that both will invade simultaneously, nor that invasion fitness of one inbreeding strategy will be independent of another. For example, if pre-adaptation or rapid selection results in the fixation of alleles underlying adaptive pre-copulatory inbreeding avoidance, then alleles underlying post-copulatory inbreeding avoidance might be precluded from invading a population in which post-copulatory choice is not already expressed because the effect that such invading alleles have on decreasing total inbreeding avoidance could be weak or negligible. To predict evolutionary outcomes, it is therefore necessary to consider evolution when one inbreeding avoidance strategy is initially rare, and the other strategy is common.

@@

<!--- Here we apply a model of female and male reproductive interactions and associated inbreeding depression in offspring viability to quantify evolution of three separate phenotypes including female tendency for polyandry, female pre-copulatory inbreeding strategy (mating strategy), and female post-copulatory inbreeding strategy (fertilisation strategy). --->

Here we use individual-based modelling and *in silico* experiments to test the hypothesis that, given severe inbreeding depression in offspring viability, evolution of post-copulatory inbreeding avoidance causes selection for evolution of polyandry. Additionally, we test the hypothesis that costs associated with polyandry and inbreeding strategies will strongly affect evolution of each such that selection for pre-copulatory inbreeding avoidance but not post-copulatory inbreeding avoidance will occur when pre-copulatory inbreeding avoidance is less costly than the cost of post-copulatory inbreeding avoidance, while selection for post-copulatory inbreeding avoidance but not pre-copulatory inbreeding avoidance will occur when pre-copulatory inbreeding avoidance is more costly than the cost of post-copulatory inbreeding avoidance. Finally, we test the hypothesis that evolution of one inbreeding strategy will be precluded if it is introduced into a population that is already fixed for a complementary inbreeding strategy that is adaptive. We thereby provide a framework for predicting evolution of pre-copulatory and post-copulatory inbreeding avoidance across reproductive systems with different relative costs and strong inbreeding depression in offspring viability.
